### 2021-02-23 paper2/3-method.tex
Lots of improvements

We optimise this loss in an adversarial fashion by playing a min-max game, in which our encoder acts as the generative component.
The adversary is an auxiliary classifier $g$, which receives $\bm{z}_u$ as input and attempts to predict the spurious variable $s$.
We denote the parameters of the adversary as $\phi$;
for the parameters of the encoder we use $\theta$, as before.
The objective from Eq~\eqref{eq:objectivetheory} is then% realised as
\begin{align}
  \min_{\theta\in\Theta} \max_{\phi\in\Phi} \mathbb{E}_{x \sim X}[\log p_\theta(x) -\lambda\mathcal{L}_c(g_\phi(f_\theta(x))); s)]
  \label{eq:objectivepractical}
\end{align}
where $\mathcal{L}_c$ is the cross-entropy between the predictions for $s$ and the provided labels.
In practice, this adversarial term is realised with a gradient reversal layer (GRL) \cite{ganin2016domain} between $\bm{z}_u$ and $g$ as is common in adversarial approaches~\cite{edwardsstorkey}.
% to fair representation learning

\subsection{The Disentanglement Dilemma} % HEADING IS JUST TO HELP ORGANIZE THOUGHTS, WILL BE DELETED LATER
The objective in Eq~\eqref{eq:objectivepractical} balances the two desiderata: predicting $y$ and being invariant to $s$.
However, in the training set $(X^\mathit{tr}, S^\mathit{tr}, Y^\mathit{tr})$,
$y$ and $s$ are so strongly correlated that removing information about $s$ inevitably removes information about $y$.
This strong correlation makes existing methods fail under this setting.
% However, this objective is complicated by the desideratum that $\bm{z}_u$ remain predictive of $y$,
% which precludes us from directly training on the target-labelled dataset $(X^\mathit{tr}, S^\mathit{tr}, Y^\mathit{tr})$,
%where $y$ and $s$ are so strongly correlated that removing information about $s$ inevitably removes information about $y$.
%We therefore need
In order to even define the right learning goal,
we require another source of information that allows us to disentangle $s$ and $y$.
For this, we assume the existence of another set of samples that follow a similar distribution to the test set,
but whilst the sensitive attribute is available, the class labels are not.
In reality, this is not an unreasonable assumption,
as, while properly annotated data is scarce, unlabelled data can be obtained in abundance (with demographic information from census data, electoral rolls, etc.).
Previous work has also considered treated ``unlabelled data'' as still having $s$ labels \cite{wick2019unlocking}.
We are restricted only in the sense that the spurious correlations we want to sever are indicated in the features.
We call this the \emph{representative set}, consisting of $X^\mathit{rep}$ and $S^\mathit{rep}$.
It fulfils $\mathcal{I}(S^\mathit{rep}; Y^\mathit{rep}) \approx 0$
(or rather, it would, if the class labels $Y^\mathit{rep}$ were available).
%{\color{red} justify the existence of such a dataset}

@@

% This That said, we do wish to draw a distinction between null-sampling and the annihilation operation featured in .
Null-sampling resembles the \emph{annihilation} operation described in \cite{xiao2017dna}, however we note that the two serve very different roles.  Whereas the annihilation operation serves as a regulariser to prevent trivial solutions (similar to \cite{jaiswal2018unsupervised}), null-sampling is used to generate the invariant representations post-training.

\subsection{Conditional Decoding}%
\label{conddec}
\noindent We first describe a VAE-based model similar to that proposed in~\cite{madras2018learning}, before highlighting some of its shortcomings that motivate the choice of an invertible representation learner.

The model takes the form of a class conditional $\beta$-VAE \cite{higgins2017beta}, in which the decoder is conditioned on the spurious attribute.
We use $\theta_{enc}, \theta_{dec} \in \theta$ to denote the parameters of the encoder and decoder sub-networks, respectively.
Concretely, the encoder component performs the mapping $x \rightarrow{\bm{z}_u}$, while $\mathcal{B}$ is instantiated as the decoder,
$\mathcal{B} \coloneqq p_{\theta_{dec}}(x|z_u, s)$, which takes in a concatenation of the learned non-spurious latent vector $\bm{z}_u$ and a one-hot encoding of the spurious label $s$ to produce a reconstruction of the input $\hat{x}$.
Conditioning on a one-hot encoding of $s$, rather than a single value, as done in \cite{madras2018learning} is the key to visualising invariant representations in the data domain.
If $\mathcal{I}(z_u; s)$ is properly minimised, the decoder can only derive its information about $s$ from the label, thereby freeing up $\bm{z}_u$ from encoding the unwanted information while still allowing for reconstruction of the input.
Thus, by feeding a zero-vector to the decoder we achieve $\hat{x} \perp s$.
The full learning objective for the cVAE is given as
\begin{align}
\begin{split}
    \mathcal{L}_{\mathrm{cVAE}} =& \mathbb{E}_{q_{\theta_{enc}}(z_u, b|x)}[\log p_{\theta_{dec}}(x|z, b) - \log p_{\theta_{dec}}(s|z_u)] \\ 
    &- \beta D_{KL}(q_{\theta_{enc}}(z_u |x) \| p(z_u))
\end{split}
\end{align}
where $\beta$ is a hyperparameter that determines the trade-off between reconstruction accuracy and independence constraints,
and $p(\bm{z}_u)$ is the prior imposed on the variational posterior.
For all our experiments, $p(\bm{z}_u)$ is realised as an Isotropic Gaussian.
Fig.~\ref{fig:cvae_diagram} summarises the procedure as a diagram.

While we show this setup can indeed work for simple problems, as~\cite{madras2018learning} before us have, we show that it lacks scalability due to disagreement between the components of the loss.
Since information about $s$ is only available to the decoder as a binary encoding,
if the relationship between $s$ and $x$ is highly non-linear and cannot be summarised by a simple on/off mechanism, as is the case if $s$ is an attribute such as gender,
off-loading information to the decoder by conditioning is no longer possible. As a result, $\bm{z}_u$ is forced to carry information about $s$ in order to minimise the reconstruction error.

The obvious solution to this is to allow the encoder to store information about $s$ in a partition of the latent space as in  \cite{creager2019flexibly}.  However, we question whether an autoencoder is the best choice for this setup, with the view that an invertible model is the better tool for the task. Using an invertible model has several guarantees, namely complete information-preservation and freedom from a reconstruction loss, the importance of which we elaborate on below.

\subsection{Conditional Flow}\label{cflow}
\subsubsection{Invertible Neural Networks}
Invertible neural networks are a class of neural network architecture characterised by a bijective mapping between their inputs and output \cite{Dinh2014}. The transformations are designed such that their inverses and Jacobians are efficiently computable.
These flow-based models permit \emph{exact} likelihood estimation \cite{normflows2015} through the warping of a base density with a series of invertible transformations and computing the resulting, highly multi-modal, but still normalised, density, using the change of variable theorem:
% Flow-GAN \cite{grover2018flowgan} combines the \emph{exact} log-likelihood estimation of the invertible network with the adversarial training of a GAN.
\begin{align}
\begin{split}
  \log p(x) &= \log p(z) + 
   \sum \log \left| \det\left( \frac{\diff h_i}{ h_{i-1}}\right) \right|, %\\
  \quad p(z) = \mathcal{N}(z; 0, \mathbb{I})
  \label{eq:changeofvariables}
\end{split}
\end{align}
where $h_i$ refers to the outputs of the layers of the network and $p(z)$ is the base density, specifically an Isotropic Gaussian in our case.
Training of the invertible neural network is then reduced to maximising $\log p(x)$ over the training set,
i.e.\ maximising the probability the network assigns to samples in the training set.

@@

\subsubsection{The Benefits of Bijectivity}
Using an invertible network to generate our encoding, $\bm{z}_u$, carries a number of advantages over other approaches.
Ordinarily, the main benefit of flow-based models is that they permit exact density estimation. 
However, since we are not interested in sampling from the model's distribution, in our case the likelihood term serves as a regulariser, as it does for  \cite{JacSmeOya18}. 
Critically, this forces the mean of each latent dimension to zero enabling null-sampling. 
The invertible property of the network guarantees the preservation of all information relevant to $y$ which is independent of $s$, regardless of how it is allocated in the output space.
Secondly, we conjecture that the encodings are more robust to out-of-distribution data.
Whereas an autoencoder could map a previously seen input and a previously unseen input to the same representation,
an invertible network sidesteps this due to the network's bijective property, ensuring all relevant information is stored somewhere. This opens up the possibility of transfer learning between datasets with a similar manifestation of $s$, as we demonstrate in the Appendix~\ref{sec:transfer-learning}.

>>>>

We optimise this loss in an adversarial fashion by playing a min-max game, in which our encoder acts as the generative component.
The adversary is an auxiliary classifier $g$, which receives $\bm{z}_u$ as input and attempts to predict the spurious variable $s$.
We denote the parameters of the adversary as $\phi$;
for the parameters of the encoder we use $\theta$, as before.
The objective from~\eqref{eq:objectivetheory} is then% realised as
\begin{align}
  \min_{\theta\in\Theta} \max_{\phi\in\Phi} \mathbb{E}_{x \sim X}[\log p_\theta(x) -\lambda\mathcal{L}_c(g_\phi(f_\theta(x))); s)]
  \label{eq:objectivepractical}
\end{align}
where $\mathcal{L}_c$ is the cross-entropy between the predictions for $s$ and the provided labels.
In practice, this adversarial term is realised with a gradient reversal layer (GRL) \citep{ganin2016domain} between $\bm{z}_u$ and $g$ as is common in adversarial approaches~\citep{edwards2016censoring}.
% to fair representation learning

\subsection{The Disentanglement Dilemma} % HEADING IS JUST TO HELP ORGANIZE THOUGHTS, WILL BE DELETED LATER
The objective in~\eqref{eq:objectivepractical} balances the two desiderata: predicting $y$ and being invariant to $s$.
However, in the training set $(X^\mathit{tr}, S^\mathit{tr}, Y^\mathit{tr})$,
$y$ and $s$ are so strongly correlated that removing information about $s$ inevitably removes information about $y$.
This strong correlation makes existing methods fail under this setting.
% However, this objective is complicated by the desideratum that $\bm{z}_u$ remain predictive of $y$,
% which precludes us from directly training on the target-labelled dataset $(X^\mathit{tr}, S^\mathit{tr}, Y^\mathit{tr})$,
%where $y$ and $s$ are so strongly correlated that removing information about $s$ inevitably removes information about $y$.
%We therefore need
In order to even define the right learning goal,
we require another source of information that allows us to disentangle $s$ and $y$.
For this, we assume the existence of another set of samples that follow a similar distribution to the test set,
but whilst the sensitive attribute is available, the class labels are not.
In reality, this is not an unreasonable assumption,
as, while properly annotated data is scarce, unlabelled data can be obtained in abundance (with demographic information from census data, electoral rolls, etc.).
Previous work has also considered treated ``unlabelled data'' as still having $s$ labels~\citep{wick2019unlocking}.
We are restricted only in the sense that the spurious correlations we want to sever are indicated in the features.
We call this the \emph{representative set}, consisting of $X^\mathit{rep}$ and $S^\mathit{rep}$.
It fulfils $\mathcal{I}(S^\mathit{rep}; Y^\mathit{rep}) \approx 0$
(or rather, it would, if the class labels $Y^\mathit{rep}$ were available).
%{\color{red} justify the existence of such a dataset}

@@

% This That said, we do wish to draw a distinction between null-sampling and the annihilation operation featured in .
Null-sampling resembles the \emph{annihilation} operation described in \citet{xiao2017dna}, however we note that the two serve very different roles.  Whereas the annihilation operation serves as a regulariser to prevent trivial solutions (similar to \cite{jaiswal2018unsupervised}), null-sampling is used to generate the invariant representations post-training.

\subsection{Conditional Decoding}%
\label{conddec}
\noindent We first describe a VAE-based model similar to that proposed in~\citet{madras2018learning}, before highlighting some of its shortcomings that motivate the choice of an invertible representation learner.

The model takes the form of a class conditional $\beta$-VAE \citep{higgins2017beta}, in which the decoder is conditioned on the spurious attribute.
We use $\theta_{enc}, \theta_{dec} \in \theta$ to denote the parameters of the encoder and decoder sub-networks, respectively.
Concretely, the encoder component performs the mapping $x \rightarrow{\bm{z}_u}$, while $\mathcal{B}$ is instantiated as the decoder,
$\mathcal{B} \coloneqq p_{\theta_{dec}}(x|z_u, s)$, which takes in a concatenation of the learned non-spurious latent vector $\bm{z}_u$ and a one-hot encoding of the spurious label $s$ to produce a reconstruction of the input $\hat{x}$.
Conditioning on a one-hot encoding of $s$, rather than a single value, as done in \citet{madras2018learning} is the key to visualising invariant representations in the data domain.
If $\mathcal{I}(z_u; s)$ is properly minimised, the decoder can only derive its information about $s$ from the label, thereby freeing up $\bm{z}_u$ from encoding the unwanted information while still allowing for reconstruction of the input.
Thus, by feeding a zero-vector to the decoder we achieve $\hat{x} \perp s$.
The full learning objective for the cVAE is given as
\begin{align}
\begin{split}
    \mathcal{L}_{\mathrm{cVAE}} =& \mathbb{E}_{q_{\theta_{enc}}(z_u, b|x)}[\log p_{\theta_{dec}}(x|z, b) - \log p_{\theta_{dec}}(s|z_u)] \\ 
    &- \beta D_{KL}(q_{\theta_{enc}}(z_u |x) \| p(z_u))
\end{split}
\end{align}
where $\beta$ is a hyperparameter that determines the trade-off between reconstruction accuracy and independence constraints,
and $p(\bm{z}_u)$ is the prior imposed on the variational posterior.
For all our experiments, $p(\bm{z}_u)$ is realised as an Isotropic Gaussian.
Fig.~\ref{fig:cvae_diagram} summarises the procedure as a diagram.

While we show this setup can indeed work for simple problems, as~\citet{madras2018learning} before us have, we show that it lacks scalability due to disagreement between the components of the loss.
Since information about $s$ is only available to the decoder as a binary encoding,
if the relationship between $s$ and $x$ is highly non-linear and cannot be summarised by a simple on/off mechanism, as is the case if $s$ is an attribute such as gender,
off-loading information to the decoder by conditioning is no longer possible. As a result, $\bm{z}_u$ is forced to carry information about $s$ in order to minimise the reconstruction error.

The obvious solution to this is to allow the encoder to store information about $s$ in a partition of the latent space as in  \citet{creager2019flexibly}.  However, we question whether an autoencoder is the best choice for this setup, with the view that an invertible model is the better tool for the task. Using an invertible model has several guarantees, namely complete information-preservation and freedom from a reconstruction loss, the importance of which we elaborate on below.

\subsection{Conditional Flow}\label{cflow}
\subsubsection{Invertible Neural Networks}
Invertible neural networks are a class of neural network architecture characterised by a bijective mapping between their inputs and output \citep{Dinh2014}. The transformations are designed such that their inverses and Jacobians are efficiently computable.
These flow-based models permit \emph{exact} likelihood estimation \citep{normflows2015} through the warping of a base density with a series of invertible transformations and computing the resulting, highly multi-modal, but still normalised, density, using the change of variable theorem:
% Flow-GAN \cite{grover2018flowgan} combines the \emph{exact} log-likelihood estimation of the invertible network with the adversarial training of a GAN.
\begin{align}
\begin{split}
  \log p(x) &= \log p(z) + 
   \sum \log \left| \det\left( \frac{\diff h_i}{ h_{i-1}}\right) \right|, %\\
  \quad p(z) = \mathcal{N}(z; 0, \mathbb{I})
  \label{eq:changeofvariables}
\end{split}
\end{align}
where $h_i$ refers to the outputs of the layers of the network and $p(z)$ is the base density, specifically an Isotropic Gaussian in our case.
Training of the invertible neural network is then reduced to maximising $\log p(x)$ over the training set,
i.e.\ maximising the probability the network assigns to samples in the training set.

@@

\subsubsection{The Benefits of Bijectivity}
Using an invertible network to generate our encoding, $\bm{z}_u$, carries a number of advantages over other approaches.
Ordinarily, the main benefit of flow-based models is that they permit exact density estimation. 
However, since we are not interested in sampling from the model's distribution, in our case the likelihood term serves as a regulariser, as it does for  \citet{JacSmeOya18}. 
Critically, this forces the mean of each latent dimension to zero enabling null-sampling. 
The invertible property of the network guarantees the preservation of all information relevant to $y$ which is independent of $s$, regardless of how it is allocated in the output space.
Secondly, we conjecture that the encodings are more robust to out-of-distribution data.
Whereas an autoencoder could map a previously seen input and a previously unseen input to the same representation,
an invertible network sidesteps this due to the network's bijective property, ensuring all relevant information is stored somewhere. This opens up the possibility of transfer learning between datasets with a similar manifestation of $s$, as we demonstrate in the Appendix~\ref{sec:transfer-learning}.

### 2021-02-23 paper2/4-experiments.tex
Lots of improvements

\section{Experiments}
\noindent We present experiments to demonstrate that the null-sampled representations are in fact invariant to $s$
while still allowing a classifier to predict $y$ from them.
We run our cVAE and cFlow models on the coloured MNIST (cMNIST) and CelebA dataset,
which we artificially bias, first describing the sampling procedure we follow to do so for non-synthetic datasets.
As baselines we have the model of~\cite{ln2l} (Ln2L) and the same CNN used to evaluate the cFlow and cVAE models
but with the unmodified images as input (CNN).
For the cFlow model we adopt a Glow-like architecture~\cite{KinDha18},
while both subnetworks of the cVAE model comprise gated convolutions~\cite{van2016conditional}, where the encoding size is $256$.
For cMNIST, we construct the Ln2L baseline according to its original description, for CelebA,
we treat it as an augmentation of the baseline CNN's objective function.
%More
Detailed information regarding model architectures can be found in Appendix~\ref{sec:architectures} and~\ref{sec:optimisation-details}.
% \begin{figure*}[!htb]
%   \centering
%   %
%   \subfloat[CelebA]{\scalebox{0.495}{\includegraphics[width=\textwidth]{./results/celeba/celeba_acc_DP.pdf}}
%   \label{fig:celeba_chart}
%   }%
%   %\hfill
%   \subfloat[UCI Adult]{\scalebox{0.495}{\includegraphics[width=\textwidth]{./Figures/nosinn_celeba_multiplot_all_landscape_Smiling.pdf}}
%   \label{fig:adult_chart}
%   }%
%   \caption{
%     Performance of our approach on the CelebA and UCI Adult datasets.
%     (Left of each) We show the accuracy of a downstream classifier trained on representations learned by our model.
%     (Right of each) We show the Demographic Parity of a model trained on our representations (the lower the better).
%   }\label{fig:chart}
% \end{figure*}
\begin{figure}[tb]
    \centering
    \includegraphics[width=0.7\textwidth]{./Figures/nosinn_celeba.pdf}
    \caption{
        Performance of our model for different targets (mixing factor $\eta=0$).
        Left: \emph{Smiling} as target, right: \emph{high cheekbones}.
        \emph{DP diff} measures fairness with respect to demographic parity.
        A perfectly fair model has a \emph{DP diff} of 0.
    }%
    \label{fig:celeba-targets}
\end{figure}

@@

% \begin{figure*}[!htb]
%   \centering
%   \caption{Performance of our methods (cVAE and cFlow) on the tabular UCI Adult Dataset. This is a common benchmark in fair machine learning literature. Each model depicts the result of a ownstream classifier trained on the embeddings produced by one of the models. (Left) Accuracy of a downstream classifier. (Middle) Demographic Parity of a downstream classifier. (Right) Equality of Opportunity of a downstream classifier.}\label{fig:adult_results}
% \end{figure*}
\subsubsection{Synthesising Dataset Bias}
For our experiments, we require a training set that exhibits a strong spurious correlation, together with a test set that does not.
For cMNIST, this is easily satisfied as we have complete control over the data generation process.
For CelebA and  UCI Adult, on the other hand,
we have to generate the split from the existing data.
To this end, we first set aside a randomly selected portion of the dataset from which to sample the biased dataset
The portion itself is then split further into two parts:
one in which $(s=-1 \land y=-1) \lor (s=+1 \land y=+1)$ holds true for all samples, call this part $\mathcal{D}_{eq}$,
and the other part, call it $\mathcal{D}_{opp}$, which contains the remaining samples.
To investigate the behaviour at different levels of correlation,
we mix these two subsets according to a mixing factor $\eta$.
For $\eta \leq \tfrac{1}{2}$, we combine (all of) $\mathcal{D}_{eq}$
with a fraction of $2\eta$ from $\mathcal{D}_{opp}$.
For $\eta > \tfrac{1}{2}$, we combine (all of) $\mathcal{D}_{opp}$
and a fraction of $2(1 -\eta)$ from $\mathcal{D}_{eq}$.
Thus, for $\eta=0$, the biased dataset is just $\mathcal{D}_{eq}$,
for $\eta=1$ it is just $\mathcal{D}_{opp}$
and for $\eta=\tfrac{1}{2}$ the biased dataset is an ordinary subset of the whole data. The test set is simply the data remaining from the initial split.

\subsubsection{Evaluation protocol}
We evaluate our results in terms of accuracy and fairness.
A model that perfectly decouples its predictions from $s$ will achieve near-uniform accuracy across all biasing-levels.
For binary $s$/$y$ we quantify the fairness of a classifier's predictions using \emph{demographic parity} (DP): the  absolute difference in the probability  of a positive prediction for each sensitive group.

@@

\subsubsection{cMNIST}\label{ssec:cmnist}
The coloured MNIST (cMNIST) dataset is a variant of the MNIST dataset in which the digits are coloured.
In the training set, the colours have a one-to-one correspondence with the digit class.
In the test set (and the representative set), colours are assigned randomly.
The colours are drawn from Gaussians with 10 different means.
We follow the colourisation procedure outlined by~\cite{ln2l}, with the mean colour values selected so as to be maximally dispersed.
The full list of such values can be found in Appendix~\ref{sec:color-details}.
We produce multiple variants of the cMNIST dataset corresponding to different standard deviations $\sigma$ for the colour sampling:
$\sigma \in \{0.00, 0.01, ..., 0.05 \}$.
% The images are symmetrically zero-padded to a size of $32 \times 32$ to allow for an additional downsampling stage in the cFlow model.
% No additional data augmentation is used.
% 24,000 of the 36,000 training samples were set aside for representative, the remaining 10,000 for training the downstream classifier.
% We continue to use 10,000 samples usually reserved for testing on vanilla MNIST for that same purpose.

For this specific dataset, we can establish an additional baseline by simply grey-scaling the dataset
% Since the data-generation process is known, we can establish a baseline an additional by following the simple heuristic of grey-scaling the dataset
which only leaves the luminosity as spurious information.
%The grey-scaling operation does not completely eliminate the bias in the dataset, it only mitigates it by reducing the number of unique values for $s$ it is still exhibited residually in the luminosity of the digits.
We also evaluate the model, with all the associated hyperparameters, from~\cite{ln2l}.
The only difference between the setups is the dataset creation, including the range of $\sigma$ values we consider.
Our versions of the dataset, on the whole, exhibit much stronger colour bias, to the point of the mapping the digit's colour and class being bijective.
Fig.~\ref{fig:cmnist_chart} shows that the model significantly underperforms even the na\"ive baseline, aside from at $\sigma = 0$, where they are on par.

@@

\subsubsection{CelebA}
To evaluate the effectiveness of our framework on real-world image data we use the CelebA dataset~\cite{liu2015faceattributes}, consisting of 202,599 celebrity images.
These images are annotated with various binary physical  attributes, including ``gender'', ``hair color'', ``young'', etc, from which we  select our sensitive and target attributes.
The images are centre cropped and resized to $64\times64$, as is standard practice.
For our experiments, we designate ``gender'' as the sensitive attribute,
and ``smiling'' and ``high cheekbones'' as target attributes.
We chose gender as the sensitive attribute as it a common sensitive attribute in the fairness literature.
For the target attributes, we chose attributes that are harder to learn than gender and which do not correlate too strongly with gender in the dataset
(``wearing lipstick'' for example being an attribute too closely correlated with gender).
% The sampling and evaluation procedure is identical to that conducted for the UCI Adult dataset.
% We evaluate the performance of all models for mixing factors ($\eta$) of value $\{0, 0.1, ..., 1\}$. 
The model is trained on the representative set (normal subset of CelebA)
and is then used to encode the artificially biased training set and the test set.
The results for the most strongly biased training set ($\eta=0$) can be found in Fig.~\ref{fig:celeba-targets}.
Our method outperforms the baselines in accuracy and fairness.

>>>>

\section{Experiments}
\noindent We present experiments to demonstrate that the null-sampled representations are in fact invariant to $s$
while still allowing a classifier to predict $y$ from them.
We run our cVAE and cFlow models on the coloured MNIST (cMNIST) and CelebA dataset,
which we artificially bias, first describing the sampling procedure we follow to do so for non-synthetic datasets.
As baselines we have the model of~\citet{kim2019learning} (Ln2L) and the same CNN used to evaluate the cFlow and cVAE models
but with the unmodified images as input (CNN).
For the cFlow model we adopt a Glow-like architecture~\citep{KinDha18},
while both subnetworks of the cVAE model comprise gated convolutions~\citep{van2016conditional}, where the encoding size is $256$.
For cMNIST, we construct the Ln2L baseline according to its original description, for CelebA,
we treat it as an augmentation of the baseline CNN's objective function.
%More
Detailed information regarding model architectures can be found in Appendix~\ref{sec:architectures} and~\ref{sec:optimisation-details}.%
\footnote{The code can be found at \url{https://github.com/predictive-analytics-lab/nifr}.}
% \begin{figure*}[!htb]
%   \centering
%   %
%   \subfloat[CelebA]{\scalebox{0.495}{\includegraphics[width=\textwidth]{./results/celeba/celeba_acc_DP.pdf}}
%   \label{fig:celeba_chart}
%   }%
%   %\hfill
%   \subfloat[UCI Adult]{\scalebox{0.495}{\includegraphics[width=\textwidth]{./Figures/nosinn_celeba_multiplot_all_landscape_Smiling.pdf}}
%   \label{fig:adult_chart}
%   }%
%   \caption{
%     Performance of our approach on the CelebA and UCI Adult datasets.
%     (Left of each) We show the accuracy of a downstream classifier trained on representations learned by our model.
%     (Right of each) We show the Demographic Parity of a model trained on our representations (the lower the better).
%   }\label{fig:chart}
% \end{figure*}
\begin{figure}[tb]
    \centering
    \includegraphics[width=0.7\textwidth]{./Figures/nosinn_celeba.pdf}
    \caption{
        Performance of our model for different targets (mixing factor $\eta=0$).
        Left: \emph{Smiling} as target, right: \emph{high cheekbones}.
        \emph{DP diff} measures fairness with respect to demographic parity.
        A perfectly fair model has a \emph{DP diff} of 0.
    }%
    \label{fig:celeba-targets}
\end{figure}

@@

% \begin{figure*}[!htb]
%   \centering
%   \caption{Performance of our methods (cVAE and cFlow) on the tabular UCI Adult Dataset. This is a common benchmark in fair machine learning literature. Each model depicts the result of a ownstream classifier trained on the embeddings produced by one of the models. (Left) Accuracy of a downstream classifier. (Middle) Demographic Parity of a downstream classifier. (Right) Equality of Opportunity of a downstream classifier.}\label{fig:adult_results}
% \end{figure*}
\subsection{Synthesising Dataset Bias}
For our experiments, we require a training set that exhibits a strong spurious correlation, together with a test set that does not.
For cMNIST, this is easily satisfied as we have complete control over the data generation process.
For CelebA and  UCI Adult, on the other hand,
we have to generate the split from the existing data.
To this end, we first set aside a randomly selected portion of the dataset from which to sample the biased dataset
The portion itself is then split further into two parts:
one in which $(s=-1 \land y=-1) \lor (s=+1 \land y=+1)$ holds true for all samples, call this part $\mathcal{D}_{eq}$,
and the other part, call it $\mathcal{D}_{opp}$, which contains the remaining samples.
To investigate the behaviour at different levels of correlation,
we mix these two subsets according to a mixing factor $\eta$.
For $\eta \leq \tfrac{1}{2}$, we combine (all of) $\mathcal{D}_{eq}$
with a fraction of $2\eta$ from $\mathcal{D}_{opp}$.
For $\eta > \tfrac{1}{2}$, we combine (all of) $\mathcal{D}_{opp}$
and a fraction of $2(1 -\eta)$ from $\mathcal{D}_{eq}$.
Thus, for $\eta=0$, the biased dataset is just $\mathcal{D}_{eq}$,
for $\eta=1$ it is just $\mathcal{D}_{opp}$
and for $\eta=\tfrac{1}{2}$ the biased dataset is an ordinary subset of the whole data. The test set is simply the data remaining from the initial split.

\subsection{Evaluation protocol}
We evaluate our results in terms of accuracy and fairness.
A model that perfectly decouples its predictions from $s$ will achieve near-uniform accuracy across all biasing-levels.
For binary $s$/$y$ we quantify the fairness of a classifier's predictions using \emph{demographic parity} (DP): the  absolute difference in the probability  of a positive prediction for each sensitive group.

@@

\subsubsection{cMNIST}\label{ssec:cmnist}
The coloured MNIST (cMNIST) dataset is a variant of the MNIST dataset in which the digits are coloured.
In the training set, the colours have a one-to-one correspondence with the digit class.
In the test set (and the representative set), colours are assigned randomly.
The colours are drawn from Gaussians with 10 different means.
We follow the colourisation procedure outlined by~\citet{kim2019learning}, with the mean colour values selected so as to be maximally dispersed.
The full list of such values can be found in Appendix~\ref{sec:color-details}.
We produce multiple variants of the cMNIST dataset corresponding to different standard deviations $\sigma$ for the colour sampling:
$\sigma \in \{0.00, 0.01, ..., 0.05 \}$.
% The images are symmetrically zero-padded to a size of $32 \times 32$ to allow for an additional downsampling stage in the cFlow model.
% No additional data augmentation is used.
% 24,000 of the 36,000 training samples were set aside for representative, the remaining 10,000 for training the downstream classifier.
% We continue to use 10,000 samples usually reserved for testing on vanilla MNIST for that same purpose.

For this specific dataset, we can establish an additional baseline by simply grey-scaling the dataset
% Since the data-generation process is known, we can establish a baseline an additional by following the simple heuristic of grey-scaling the dataset
which only leaves the luminosity as spurious information.
%The grey-scaling operation does not completely eliminate the bias in the dataset, it only mitigates it by reducing the number of unique values for $s$ it is still exhibited residually in the luminosity of the digits.
We also evaluate the model, with all the associated hyperparameters, from~\citet{kim2019learning}.
The only difference between the setups is the dataset creation, including the range of $\sigma$ values we consider.
Our versions of the dataset, on the whole, exhibit much stronger colour bias, to the point of the mapping the digit's colour and class being bijective.
Fig.~\ref{fig:cmnist_chart} shows that the model significantly underperforms even the na\"ive baseline, aside from at $\sigma = 0$, where they are on par.

@@

\subsubsection{CelebA}
To evaluate the effectiveness of our framework on real-world image data we use the CelebA dataset~\citep{liu2015faceattributes}, consisting of 202,599 celebrity images.
These images are annotated with various binary physical  attributes, including ``gender'', ``hair color'', ``young'', etc, from which we  select our sensitive and target attributes.
The images are centre cropped and resized to $64\times64$, as is standard practice.
For our experiments, we designate ``gender'' as the sensitive attribute,
and ``smiling'' and ``high cheekbones'' as target attributes.
We chose gender as the sensitive attribute as it a common sensitive attribute in the fairness literature.
For the target attributes, we chose attributes that are harder to learn than gender and which do not correlate too strongly with gender in the dataset
(``wearing lipstick'' for example being an attribute too closely correlated with gender).
% The sampling and evaluation procedure is identical to that conducted for the UCI Adult dataset.
% We evaluate the performance of all models for mixing factors ($\eta$) of value $\{0, 0.1, ..., 1\}$. 
The model is trained on the representative set (normal subset of CelebA)
and is then used to encode the artificially biased training set and the test set.
The results for the most strongly biased training set ($\eta=0$) can be found in Fig.~\ref{fig:celeba-targets}.
Our method outperforms the baselines in accuracy and fairness.
